---
title: "Reproducible Research Project 1"
author: "Avantika Deb"
output: md_document
---
## Introduction

A large amount of data can now be collected about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, Jawbone Up etc. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data

The data for this assignment can be downloaded using the following URL:- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

The variables included in this dataset are:

1. steps: Number of steps taking in a 5-minute interval
2. date: The date on which the measurement was taken in YYYY-MM-DD format
3. interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and processing the data
```{r, echo=TRUE, message=FALSE}
library(ggplot2)
library(dplyr)
```
```{r, echo=TRUE}
url<- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = paste0(getwd(), '/zipfile.zip'))
 unzip("zipfile.zip")
 
 activitydata<- read.csv('activity.csv')
 activitydata$date <- as.Date(activitydata$date, '%Y-%m-%d')
```
## Mean total number of steps taken per day

**Calculate the total number of steps taken per day**
```{r, echo=TRUE}
totalsteps <- tapply(activitydata$steps, activitydata$date, sum, na.rm=T)
head(totalsteps)
```
**Histogram of the total number of steps taken each day**
```{r, echo=TRUE}
hist(totalsteps, xlab="Total steps per day")
```

**Mean and median of the total number of steps taken per day**
```{r, echo=TRUE}
summary(totalsteps)
```

## Average daily activity pattern

**Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**
```{r, echo=TRUE}
averagesteps <- activitydata %>% filter(!is.na(steps)) %>% group_by(interval) %>% summarize(steps = mean(steps))
ggplot(activitydata, aes(x=interval, y=steps)) + geom_line(size=1)+ labs(x = "Interval", y = "Average steps per day")
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r, echo=TRUE}
maxsteps <- averagesteps[which.max(averagesteps$steps), ][[1]]
maxsteps
```
## Imputing missing values

**The total number of missing values in the dataset**
```{r, echo=TRUE}
missing<- is.na(activitydata)
sum(missing)
```
**We will fill in the NA values with mean number of steps**
**New dataset that is equal to the original dataset but with the missing data filled in**

```{r, echo=TRUE}
finalsteps <- activitydata$steps
finalsteps[is.na(finalsteps)] <- round(mean(activitydata$steps, na.rm = T))
finalsteps <- as.numeric(finalsteps)
finaldata <- cbind(finalsteps, activitydata$date, activitydata$interval)
colnames(finaldata) <- colnames(activitydata)
finaldata<- as.data.frame(finaldata)
```

**Histogram of the total number of steps taken each day**
```{r, echo=TRUE}
finaltotalsteps <- tapply(finaldata$steps, finaldata$date, sum)
hist(finaltotalsteps, xlab="Total steps per day")
```

**Mean and median total number of steps taken per day**

```{r, echo=TRUE}
summary(finaltotalsteps)
```
The mean and median total number of steps taken per day increase slightly with the imputation of the missing values.

## Are there differences in activity patterns between weekdays and weekends?


**New factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.**

```{r, echo=TRUE}
finaldata$date <- as.POSIXct.Date(finaldata$date)
weekd <-c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
finaldata$weekday <- factor((weekdays(finaldata$date) %in% weekd), levels = c(FALSE, TRUE), labels = c('weekend', 'weekday'))
```

**Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)**

```{r, echo=TRUE}
finalavgsteps<- finaldata %>% group_by(interval, weekday) %>% summarise(steps = mean(steps))
ggplot(finalavgsteps, aes(x=interval, y=steps)) +
        geom_line(size=1) + 
        facet_grid(weekday~.)
```
